Do educational institutions explain SES gaps?: an interntational comparison of the SES achievement gap from 2000-2015

### 2017-10-09 paper/paper.tex
Spell-checked the paper

\begin{abstract}
The literature on achievement inequality has only recently started studying the evolution of the socio-economic achievement gap in cognitive abilities and it has unconvered a wide variety of results. The main finding comes from American research and it shows that the gap has been widening by about 40-50\% in the last 30 year. Other international comparisons show that there's important variation among developed countries, but the research on this area is scarce. This paper looks to study whether there are specific patterns in the evolution of the gap in a comparative perspective. We find that there is considerable variation in the way in which the gap is evolving, with the U.S and Germany closing at a rapid pace and France widening at a similar rate. We find that this gap is strongly correlated with a countrys overall performance and simulations show that switching to less tracking can help decrease a countries SES gap by about 20\%. We also find that tracking and vocational enrollment explain over one third of the evolution of the achievement gap and show that the relationship is conditioned by a strong interaction: low tracking is associated with a small achievement gap, but high levels of tracking coped with high vocational enrollment can also remedy the adversive effects of tracking.
\end{abstract}

@@

To investigate the above mentioned questions we will be using the Programme for International Student Assessment (PISA). PISA is a survey carried out every three years that aims to evaluate education systems by testing the skills and knowledge of 15-year-old students. Currrently, PISA has six waves starting in 2000 up until 2015, where recently, over half a million students were tested in mathematics, literacy and science in over 70 developed/developing countries.

PISA collects data through a two-stage stratified sampling design. With the help of official governments, PISA randomly chooses 150 schools in each country, where they then randomly pick thirty 15 year olds to undertake the two hour tests. The sample size for PISA 2000 is 127,388, 276,165 for PISA 2003, 398,750 for PISA 2006, 515,958 for PISA 2009, 480,810 for PISA 2012, and 519,334 for PISA 2015. Together with the subject tests, PISA collects personal information from students, their families and their school environment (including teacher surveys), that serves as relevent background information that can be matched to the students performance. With the recent inclusion of PISA 2015, these six waves make up a time-series analysis of 15 years, enough to visualize changes in the structure of an educational system. None of the literature cited so far has used the last PISA wave, which was recently released in December 2016.

@@

We start by look at the achievement for some countries in figure 2. We plot only mathematics for each country and also a quadratic trend spline for \emph{both mathematics and literacy} pooled. Some countries have increased their achievement very strongly. For example, France, Austria and surprisingly Sweden have very steep slopes. France increased the gap by roughly 0.9 standard deviations, Austria by 0.6 and Sweden by 0.6. This pattern happens similarly for literacy. For example, France has an increase of 0.6. For such a short period of time, the magnitud of these increases are reasonably big.

@@

On the other hand, we do have some countries which show a decrease in the SES achievement gap. Poland decreased by about -0.4 and Denmark by -0.2. However, the most notable cases are the United States and Germany. These two countries show a very high level of dispersion in the year 2000 with SES gaps of over 2 standard deviations. But in the 15-year time trend both countries reduced the gaps by -0.6 and -0.9 respectively. Their distinctively big gaps in 2000 also show up in the work of \citet{micklewright}. This corroborates the findings of \citet{reardon_portilla}, which found a decreasing gap for kindergarteners. This decline is evident also for 15 year olds, suggesting it might be more of an institutional change rather than a specific grade-level policy. Do note that the trendline is for both mathematics and literacy. If it were only for mathematics, then the trendline would probably be even more pronounced. Despite this, pooling both subjects gives even more strength to the analysis given that the results stand for both subjects together \footnote{Same graph is availabe for reading upon request, where we find pretty much the same results if not even more steep/downward slopes.}. We also plot the 80/20 and 70/30 SES gaps and find very similar patterns (see figure 13 in the appendix).

Analyzing this graph we might naively conclude that these trends are not very steep and they should not be very important to consider. However, remember that the Y axis is measured in standard deviations. Small changes are actually very huge in practical terms. Take the case of Sweden. The slope doesn't look that steep but in reality it increased the gap from 1 standard deviation in 2000 to around 1.5 standard deviation in 2015. Now that we know that, the trends of Poland, United States, France and Germany are particularly shocking.

@@

Generally speaking, we see that most countries increased their achievement gap over time. France had an average increase of about 80\% since 2000 varying down to 40\%, whereas Germany had a similar figure but decreasing. Many of the countries we saw before that didn't have a very steep slope, such as Hungary or Australia, in fact had increases of about 40\% of their gap. In contrast, we see that the U.S and Poland had also quite significant decreases of about 40\%. The benefit of presenting these estimates this way is that we can actually asses the uncertainty of each calculation, and we do find that they have wide variability. Despite this, most countries show a clear sign of either decreasing or increasing. We plot the same graph for the 80/20 (figure 14) and 70/30 (figure 15) SES gaps in the appendix and the results hold for these achievement gaps as well.

@@

Something reassuring is that mathematics and reading (not presented) follow basically the same trend across all countries \footnote{Reading had an even bigger decrease for similar countries, and a much smaller increase. The previos two graphs for reading are available upon request}. This means that the result is not an artifact of chance alone. However, it's important to disentangle where is this gap originating from. Is this because the top are improving while the bottom decreases? Or is it that the bottom is catching up? Up next we plot the same graph but show the divergent patterns between High/Low SES origins.

@@

We now pay attention to the flat summary lines. These lines should be interpreted with caution because it doesn't mean that the gap is not changing. Denmark, Belgium, Netherlands, Norway and Spain show a flat line because the gap is growing very little, if at all. On the other hand, Sweden, France, Finland, New Zealand, Austria, among other countries, show a flat line because both groups are distancing themselves at a very similar rate. These results highlight the importance of not only summarizing average achievement gaps: the source of the gap varies greatly between countries and we see how each of these patterns contibutes to the overall inequality of a country.

After analyzing the trend of each country and where it's coming from, we start to find some interesting results. We see that the United States is closing the achievement gap at a very rapid rate, but at the same time we notice that it's because the top SES group is going down faster than the bottom is catching up. In contrast, we see countries like Germany and Poland where the gap is closing as well but because the bottom group is catching up faster than the high SES group is coming down. In fact, the Polish case is the most egalitarian of all: the bottom group is catching up at a rapid pace and the top group is maintaining their high levels of performance, fostering a trully equalizing effect.

@@

We start off with Denmark. Generally speaking, we find that the 50th/10th gap is contributing more than the 90th/50th gap as the gap is bigger in mathematics. Furthermore, we find, just as in the 90th/10th SES gap, that there isn't a significant decrease \footnote{With the exception of reading in the 50th/10th SES gap which seemed to decrease from nearly 1.5 to 1 standard deviation.}. France shows a similar pattern as Denmark in that the 50th/10th SES gap is the main contributer and that the three gaps have the similar trend.

@@

The U.S shows a different pattern from all other countries. Both the 90th/50th/10th gaps seem to contribute rather similarly to the overall 90th/10th gap (although 50th/10th is marginally higher) but more interestingly, the 90th/50th gap has stagnated in the last 15 years. The top/middle groups have a relatively big gap and it hasn't changed at all. On the other hand, the steep decrease we saw for the U.S is coming exclusively from the 50th/10th gap shrinking. The reason, nonetheless, might be different from the German case, where it's probable that the bottom is catching up. Here, as in the 90th/10th gap, the middle might be 'dumbing' down closer to the low SES group. The surprising result is that the reductiong of U.S gap is coming from the shrinking of the 50/10 gap and not the 90/10 gap. These results are puzzling. Does this mean that the overall distribution of inequality in the U.S is being reduced toward the benefit of the Low SES group? The reduction is not uniform across all SES groups. We see that the 90/10 gap is closing, yet the the 90/50 gap is very stable over time. Despite this, the 50/10 is also shrinking, meaning that the 90 and 50 SES groups are shrinking down relative to the bottom while the bottom is coming up. This is something we can only speculate but the evidence points toward something like that.

@@

We see that there are more countries where High SES is increasing inequality faster than low SES is catching up (top-left panel) but there's also a fair share where low SES is catching up faster, such as the United States and Germany. Regardless of this, we see an overall negative relationship where countries in which the low SES group is catching up, the high SES group goes down. This is very interesting because it suggests that there's a trend for countries to have the bottom SES groups are climbing upwards, a trait of meritocracy. However, this means that the top group comes down, rather than stagnate or increase. Equally important, we see very few countries where low SES are decreasing (so going even further down) and High SES is also decresing (bottom left panel). On the other hand, we do see more countries where the both groups are increasing (top right panel). We plot all remaining countries in the back, together with a trendline and we see that the relation holds for all PISA countries.

To sum up, we find no evidence in support of our first hypothesis. We uncover that the tracking setup of a country doesn't imminently mean a certain degree of equality/inequality. Countries where the top SES group is increasing and the bottom is decreasing (top-left panel) have dramatically different tracking setups. Austria is an early tracker with several tracks, while Sweden is a late tracker with only 1 track. On the opossite panel (bottom-right), where we find countries that the bottom group is catching up while the top is coming down, we find similar results. Denmark and Germany are opposite poles in terms of tracking, while the U.S is distinctive in that it has little formal curricular stratification. We see no definitive evidence of top-to-bottom equality in all of these countries (except for Poland). We find that regardless of the different tracking setups, there's a tradeoff: either the bottom comes up and the top comes down or the top goes even further up and the bottom further down.

@@

Moving on, we now discuss if there's a tradeoff between an educational system being efficient, i.e raising the overall performance of the country, and whether that challenge compromises equality at the same time. For example, in order to improve the average score in, let's say, mathematics, tracking can improve the average score by making children in the higher trackers perform better, raising the overall performance. This, however, can come at a price. If tracking actually benefits the top but doesn't increase the performance of the bottom tracks then we see a net increase of inequality accompanied by an increase in the overall performance. Given that we have the data for the average performance for each country and the achievement gap, we can investigate this relationship in detail. We first being by plotting all countries pooled for each year in figure 8.

@@

Looking at figure 9, we see a strong correspondence between the two variables. In countries like France we see quite a functional form between both inequality and performance. As average performance plummets, the achievement gap is increasing. This pattern is also seen in Austria, Australia and Germany. Yet we also see the opposite trend. We see countries like Poland, Finland, Canada and less pronounced, Italy, where this pattern is evident. The surprising case here is the United States. We see that as the average performance has been decreasing, the average achievement gap has been decreasing as well. In fact, among the few countries we should find something that resembles this pattern should be the U.S because it doesn't have a strict institutionalized tracking system. However, the explanation behind this pattern is very speculative.

Finally, the last notable pattern is that in some countries the average performance is so high that it actually overperforms the achievement gap. Only Finland, Canada and Netherlands (three very equal countries) have their average performance in standard deviations above their level of inequality, which coincidentially, are countries that have the lower level of inequalities among the ones presented here.

These results actually confirm our second hypothesis, that there's a negative relationship between the two variables. We see countries like Poland, Canada and Germany where the pattern is just as the aggregate trend, namely that as the average performance increases, inequality seems to be reduced in parallel. Conversely, France, Austria and Australia, to name a few, portray the opposite: as average performance decreases inequality increases. But we also unconver important variation within countries in the relationship between average achievement and dispersion of achievement (have a look at the Netherlands, Italy and the U.S).

@@

Following the literature on the topic, we suspect that this evolution is in part very tightly related to the degree of stratification of the educational system. In a very simple frame-work, \citet{hanushek_woesmann_tracking} found that tracking indeed exarcebates inequality in the life course of children. Following the same strategy, we believe that tracking is also allowing the same gap in 15 year olds to grow over time, acompannied by other things. Yet, instead of concentrating only on the age of selection (as they do), we explore a more fine grained definition of tracking, which is possible through the work of \citet{bol2013}. Aside from the age at first selection into tracks, we also use the the number of tracks in the country, the percentage of the entire curriculum that is tracked and a vocational index \footnote{This index is a factor loading from a principal factor analysis of the percentage of students in upper secondary vocational education (taken from two sources, to reduce measurement error). We take all of the data from \citet{bol2013}.}.

Considering that our summarized data has very few observartions \footnote{We have per year, per country, one single inequality indicator, leaving the total sample size at 195}, we want to asses the uncertainty of our model by using a fully bayesian hierarchichal linear model. The benefits of this approach are twofold: first, we can asses the model fit much more intuitively, something very important given our sample size, and secondly, given that our countries are clustered into years, we can control for country heterogeneity within years, removing any bias in the standard errors.

@@

Given that the classical p values are inexistent in bayesian modelling, we present uncertainty intervals next to each coefficient. The first model only includes the 1 track dummy. This variable shows that countries that have one track, have on average, half a standard deviation of less inequality than other countries. This covariate alone explains about 23\% of the variance of the evolution of the 90/10 achievement gap. The second model now includes the dummy of whether the country has an age selection of 15 or more and we can see that countries with late selection into tracking have about -.38 less standard deviations in inequality. Also note that the number of tracks dummy was now reduced, suggesting these two variables are explaining similar things. This is expected given that late tracking is usually associated with fewer tracks, such as in the Scandinavian countries. The third model now includes whether the country has any degree of tracking, meaning whether the country has above 0\% of the curriculum tracked. The coding of the variable comes from visual inspection of the distribution, where some countries, such as the United States and Canada, have untracked curriculums. For this last variable we see no relationship at all, but after including the vocational index, the model is completely reversed with countries with tracked curriculum having about 0.6 more standard deviations in inequality that countries with no tracking.

@@

The biggest reductions are for both Hungary and Germany, with 46\% and 36\% respectively. France, which had a particularly steep increase in the last 15 years, would see a reduction of the 90/10 SES gap of 31\% and Austria, with a very similar tracking structure as the German case, would see their achievement gap reduced by 20\%. The very suprising result is the Dutch case, for which a switch from the 'worst' tracking setup to the 'best' case would result in an increase of the gap of 19\%.

This last result could mean two things. First and foremost, the Dutch educational system has built other mechanisms that, despite the tracking system, help children from low social origins escalate upwards. This explanation is not very trivial considering that in most mobility research, the Netherlands stands out as a very equal country relative to other countries with a similary tracking structure, like Germany \citep{shavit1993, esping2012}. Another possible explanation, and something which needs to assessed, is whether our model is predicting accurately. If we look at the plot from above, we see that for some countries, the model is a bit off from the actual level of inequality. Moreover, although we see that if all countries switched to the 'ideal' tracking setup we would see a vast reduction in inequality, we also see that each of these predictions is followed by a wide uncertainty interval.

@@

The exact mechanisms through whick the tracking setup is explaining these inequalities is still very speculative. We could argue that tracking might've introduced the inequality once it was implemented. Another possibility is that tracking, althought not changing, is currently exarcebating it every year. Another explanation, which we find strong evidence for, is that the tracking setup interacts very closely with the vocational setup. These explanations are very speculative and further research should carry the task of looking for a more concrete explanation.

@@

The literature on achievement inequality has recently started studying the evolution of the achievement gap and it has unconvered a wide variety of results. The main finding is that the U.S gap has been widening over time by about 40-50\% from 1970 until the 2000. Others find that this recent trend has reversed, at least for an age-specific group since the 2000's. From an international perspective, recent studies have shown that there is wide variability in this trend, with some countries showing decreases while others increases. This paper looks to investigate the magnitud of these gaps across selected countries, and attempts to explain why they are happening.

We start off by looking at the evolution of the achievement gap and document the same variation for the 15-year old achievement gap across countries. The United States has been closing its 90/10 gap by about 50\% with a wide interval going down to nearly 35\%. This reduction is also present in Germany, quite shockingly a country with a very institutionalized tracking system. Other countries such as France and Austria experience the contrary with persistent increases in the achievement gap. Moreover, once we disaggregate these patterns into a more fine-graned analysis, we discover that the dynamics are very different. The U.S gap is shrinking because both the 90th and 50th SES groups are actually going down while the bottom group is coming up. These results are evident because the 90/10 and 50/10 gap are shrinking at a very rapid rate, whereas the 90/50 gap is flat in the same 15 year interval. Germany, which has a very similar reduction, portrays a different explanation. All three achievement gaps are shrinking suggesting a more uniform reduction. On the other hand, for countries which are experiencing a widening of the gap, we see that it's happening at all instances of the distribution; all SES groups show a very steep increase.

@@

We complete the analysis by studying what explains the evolution of this inequality and we find that the degree of tracking of a country seems to be a possible explanation, but it shouldn't be studied without factoring in the degree of vocational enrollment. We find that if a country has a low level of tracking, vocational enrollment will not help decrease the achievement gap. Conversely, once tracking enters the picture, vocational enrollment can help unease the burden of inequality and reduce the achievement gap very strongly. In a similar vein, we use this model to peform a simulation for each country: if every country switched to little or no tracking, what would be the resulting reduction in the 90/10 achievement gap? We find that if all countries switched, we would find that on average countries would see their gap shrink by about 20\%, with countries like Hungary and Germany experiencing reductions of 46\% and 36\% respectively. We contrast the former simulation with another counterfactual scenario: what would be the average gap increase if all countries switched to a system where tracking predominates? We record an average widening of 72\%, something that suggests that increasing tracking is more than three times worse than reducing it.

We conclude by showing that tracking and vocational enrollment are playing an important role in explaining the evolution of the achievement gap. We find that these results are very suggestive that institutional features, more specifically, stratified curriculums, might be playing a very sizable role in the promotion of inequality. After having ran a battery of tests and different empirical strategies, the results are consistent and point out that we should pay attention to the source of the achievement gaps and study whether particular SES groups are contributing more to the gap than others. Aside from the results, this paper also revelead some puzzling questions that we leave up to future research to study.

>>>>

\begin{abstract}
The literature on achievement inequality has only recently started studying the evolution of the socio-economic achievement gap in cognitive abilities and it has uncovered a wide variety of results. The main finding comes from American research and it shows that the gap has been widening by about 40-50\% in the last 30 year. Other international comparisons show that there's important variation among developed countries, but the research on this area is scarce. This paper looks to study whether there are specific patterns in the evolution of the gap in a comparative perspective. We find that there is considerable variation in the way in which the gap is evolving, with the U.S and Germany closing at a rapid pace and France widening at a similar rate. We find that this gap is strongly correlated with a country's overall performance and simulations show that switching to less tracking can help decrease a countries SES gap by about 20\%. We also find that tracking and vocational enrollment explain over one third of the evolution of the achievement gap and show that the relationship is conditioned by a strong interaction: low tracking is associated with a small achievement gap, but high levels of tracking coped with high vocational enrollment can also remedy the aversive effects of tracking.
\end{abstract}

@@

To investigate the above mentioned questions we will be using the Programme for International Student Assessment (PISA). PISA is a survey carried out every three years that aims to evaluate education systems by testing the skills and knowledge of 15-year-old students. Currently, PISA has six waves starting in 2000 up until 2015, where recently, over half a million students were tested in mathematics, literacy and science in over 70 developed/developing countries.

PISA collects data through a two-stage stratified sampling design. With the help of official governments, PISA randomly chooses 150 schools in each country, where they then randomly pick thirty 15 year olds to undertake the two hour tests. The sample size for PISA 2000 is 127,388, 276,165 for PISA 2003, 398,750 for PISA 2006, 515,958 for PISA 2009, 480,810 for PISA 2012, and 519,334 for PISA 2015. Together with the subject tests, PISA collects personal information from students, their families and their school environment (including teacher surveys), that serves as relevant background information that can be matched to the students performance. With the recent inclusion of PISA 2015, these six waves make up a time-series analysis of 15 years, enough to visualize changes in the structure of an educational system. None of the literature cited so far has used the last PISA wave, which was recently released in December 2016.

@@

We start by look at the achievement for some countries in figure 2. We plot only mathematics for each country and also a quadratic trend spline for \emph{both mathematics and literacy} pooled. Some countries have increased their achievement very strongly. For example, France, Austria and surprisingly Sweden have very steep slopes. France increased the gap by roughly 0.9 standard deviations, Austria by 0.6 and Sweden by 0.6. This pattern happens similarly for literacy. For example, France has an increase of 0.6. For such a short period of time, the magnitude of these increases are reasonably big.

@@

On the other hand, we do have some countries which show a decrease in the SES achievement gap. Poland decreased by about -0.4 and Denmark by -0.2. However, the most notable cases are the United States and Germany. These two countries show a very high level of dispersion in the year 2000 with SES gaps of over 2 standard deviations. But in the 15-year time trend both countries reduced the gaps by -0.6 and -0.9 respectively. Their distinctively big gaps in 2000 also show up in the work of \citet{micklewright}. This corroborates the findings of \citet{reardon_portilla}, which found a decreasing gap for kindergarteners. This decline is evident also for 15 year olds, suggesting it might be more of an institutional change rather than a specific grade-level policy. Do note that the trendline is for both mathematics and literacy. If it were only for mathematics, then the trendline would probably be even more pronounced. Despite this, pooling both subjects gives even more strength to the analysis given that the results stand for both subjects together \footnote{Same graph is available for reading upon request, where we find pretty much the same results if not even more steep/downward slopes.}. We also plot the 80/20 and 70/30 SES gaps and find very similar patterns (see figure 13 in the appendix).

Analyzing this graph we might naively conclude that these trends are not very steep and they should not be very important to consider. However, remember that the Y axis is measured in standard deviations. Small changes are actually very huge in practical terms. Take the case of Sweden. The slope does not look that steep but in reality it increased the gap from 1 standard deviation in 2000 to around 1.5 standard deviation in 2015. Now that we know that, the trends of Poland, United States, France and Germany are particularly shocking.

@@

Generally speaking, we see that most countries increased their achievement gap over time. France had an average increase of about 80\% since 2000 varying down to 40\%, whereas Germany had a similar figure but decreasing. Many of the countries we saw before that did not have a very steep slope, such as Hungary or Australia, in fact had increases of about 40\% of their gap. In contrast, we see that the U.S and Poland had also quite significant decreases of about 40\%. The benefit of presenting these estimates this way is that we can actually assess the uncertainty of each calculation, and we do find that they have wide variability. Despite this, most countries show a clear sign of either decreasing or increasing. We plot the same graph for the 80/20 (figure 14) and 70/30 (figure 15) SES gaps in the appendix and the results hold for these achievement gaps as well.

@@

Something reassuring is that mathematics and reading (not presented) follow basically the same trend across all countries \footnote{Reading had an even bigger decrease for similar countries, and a much smaller increase. The previous two graphs for reading are available upon request}. This means that the result is not an artifact of chance alone. However, it's important to disentangle where is this gap originating from. Is this because the top are improving while the bottom decreases? Or is it that the bottom is catching up? Up next we plot the same graph but show the divergent patterns between High/Low SES origins.

@@

We now pay attention to the flat summary lines. These lines should be interpreted with caution because it does not mean that the gap is not changing. Denmark, Belgium, Netherlands, Norway and Spain show a flat line because the gap is growing very little, if at all. On the other hand, Sweden, France, Finland, New Zealand, Austria, among other countries, show a flat line because both groups are distancing themselves at a very similar rate. These results highlight the importance of not only summarizing average achievement gaps: the source of the gap varies greatly between countries and we see how each of these patterns contributes to the overall inequality of a country.

After analyzing the trend of each country and where it's coming from, we start to find some interesting results. We see that the United States is closing the achievement gap at a very rapid rate, but at the same time we notice that it's because the top SES group is going down faster than the bottom is catching up. In contrast, we see countries like Germany and Poland where the gap is closing as well but because the bottom group is catching up faster than the high SES group is coming down. In fact, the Polish case is the most egalitarian of all: the bottom group is catching up at a rapid pace and the top group is maintaining their high levels of performance, fostering a truly equalizing effect.

@@

We start off with Denmark. Generally speaking, we find that the 50th/10th gap is contributing more than the 90th/50th gap as the gap is bigger in mathematics. Furthermore, we find, just as in the 90th/10th SES gap, that there is not a significant decrease \footnote{With the exception of reading in the 50th/10th SES gap which seemed to decrease from nearly 1.5 to 1 standard deviation.}. France shows a similar pattern as Denmark in that the 50th/10th SES gap is the main contributer and that the three gaps have the similar trend.

@@

The U.S shows a different pattern from all other countries. Both the 90th/50th/10th gaps seem to contribute rather similarly to the overall 90th/10th gap (although 50th/10th is marginally higher) but more interestingly, the 90th/50th gap has stagnated in the last 15 years. The top/middle groups have a relatively big gap and it has not changed at all. On the other hand, the steep decrease we saw for the U.S is coming exclusively from the 50th/10th gap shrinking. The reason, nonetheless, might be different from the German case, where it's probable that the bottom is catching up. Here, as in the 90th/10th gap, the middle might be 'dumbing' down closer to the low SES group. The surprising result is that the reduction of U.S gap is coming from the shrinking of the 50/10 gap and not the 90/10 gap. These results are puzzling. Does this mean that the overall distribution of inequality in the U.S is being reduced toward the benefit of the Low SES group? The reduction is not uniform across all SES groups. We see that the 90/10 gap is closing, yet the the 90/50 gap is very stable over time. Despite this, the 50/10 is also shrinking, meaning that the 90 and 50 SES groups are shrinking down relative to the bottom while the bottom is coming up. This is something we can only speculate but the evidence points toward something like that.

@@

We see that there are more countries where High SES is increasing inequality faster than low SES is catching up (top-left panel) but there's also a fair share where low SES is catching up faster, such as the United States and Germany. Regardless of this, we see an overall negative relationship where countries in which the low SES group is catching up, the high SES group goes down. This is very interesting because it suggests that there's a trend for countries to have the bottom SES groups are climbing upwards, a trait of meritocracy. However, this means that the top group comes down, rather than stagnate or increase. Equally important, we see very few countries where low SES are decreasing (so going even further down) and High SES is also decreasing (bottom left panel). On the other hand, we do see more countries where the both groups are increasing (top right panel). We plot all remaining countries in the back, together with a trendline and we see that the relation holds for all PISA countries.

To sum up, we find no evidence in support of our first hypothesis. We uncover that the tracking setup of a country does not imminently mean a certain degree of equality/inequality. Countries where the top SES group is increasing and the bottom is decreasing (top-left panel) have dramatically different tracking setups. Austria is an early tracker with several tracks, while Sweden is a late tracker with only 1 track. On the opposite panel (bottom-right), where we find countries that the bottom group is catching up while the top is coming down, we find similar results. Denmark and Germany are opposite poles in terms of tracking, while the U.S is distinctive in that it has little formal curricular stratification. We see no definitive evidence of top-to-bottom equality in all of these countries (except for Poland). We find that regardless of the different tracking setups, there's a tradeoff: either the bottom comes up and the top comes down or the top goes even further up and the bottom further down.

@@

Moving on, we now discuss if there's a tradeoff between an educational system being efficient, i.e raising the overall performance of the country, and whether that challenge compromises equality at the same time. For example, in order to improve the average score in, let's say, mathematics, tracking can improve the average score by making children in the higher trackers perform better, raising the overall performance. This, however, can come at a price. If tracking actually benefits the top but does not increase the performance of the bottom tracks then we see a net increase of inequality accompanied by an increase in the overall performance. Given that we have the data for the average performance for each country and the achievement gap, we can investigate this relationship in detail. We first being by plotting all countries pooled for each year in figure 8.

@@

Looking at figure 9, we see a strong correspondence between the two variables. In countries like France we see quite a functional form between both inequality and performance. As average performance plummets, the achievement gap is increasing. This pattern is also seen in Austria, Australia and Germany. Yet we also see the opposite trend. We see countries like Poland, Finland, Canada and less pronounced, Italy, where this pattern is evident. The surprising case here is the United States. We see that as the average performance has been decreasing, the average achievement gap has been decreasing as well. In fact, among the few countries we should find something that resembles this pattern should be the U.S because it does not have a strict institutionalized tracking system. However, the explanation behind this pattern is very speculative.

Finally, the last notable pattern is that in some countries the average performance is so high that it actually overperforms the achievement gap. Only Finland, Canada and Netherlands (three very equal countries) have their average performance in standard deviations above their level of inequality, which coincidentally, are countries that have the lower level of inequalities among the ones presented here.

These results actually confirm our second hypothesis, that there's a negative relationship between the two variables. We see countries like Poland, Canada and Germany where the pattern is just as the aggregate trend, namely that as the average performance increases, inequality seems to be reduced in parallel. Conversely, France, Austria and Australia, to name a few, portray the opposite: as average performance decreases inequality increases. But we also uncovered important variation within countries in the relationship between average achievement and dispersion of achievement (have a look at the Netherlands, Italy and the U.S).

@@

Following the literature on the topic, we suspect that this evolution is in part very tightly related to the degree of stratification of the educational system. In a very simple frame-work, \citet{hanushek_woesmann_tracking} found that tracking indeed exacerbates inequality in the life course of children. Following the same strategy, we believe that tracking is also allowing the same gap in 15 year olds to grow over time, accompanied by other things. Yet, instead of concentrating only on the age of selection (as they do), we explore a more fine grained definition of tracking, which is possible through the work of \citet{bol2013}. Aside from the age at first selection into tracks, we also use the the number of tracks in the country, the percentage of the entire curriculum that is tracked and a vocational index \footnote{This index is a factor loading from a principal factor analysis of the percentage of students in upper secondary vocational education (taken from two sources, to reduce measurement error). We take all of the data from \citet{bol2013}.}.

Considering that our summarized data has very few observations \footnote{We have per year, per country, one single inequality indicator, leaving the total sample size at 195}, we want to assess the uncertainty of our model by using a fully bayesian hierarchical linear model. The benefits of this approach are twofold: first, we can assess the model fit much more intuitively, something very important given our sample size, and secondly, given that our countries are clustered into years, we can control for country heterogeneity within years, removing any bias in the standard errors.

@@

Given that the classical p values are inexistent in bayesian modeling, we present uncertainty intervals next to each coefficient. The first model only includes the 1 track dummy. This variable shows that countries that have one track, have on average, half a standard deviation of less inequality than other countries. This covariate alone explains about 23\% of the variance of the evolution of the 90/10 achievement gap. The second model now includes the dummy of whether the country has an age selection of 15 or more and we can see that countries with late selection into tracking have about -.38 less standard deviations in inequality. Also note that the number of tracks dummy was now reduced, suggesting these two variables are explaining similar things. This is expected given that late tracking is usually associated with fewer tracks, such as in the Scandinavian countries. The third model now includes whether the country has any degree of tracking, meaning whether the country has above 0\% of the curriculum tracked. The coding of the variable comes from visual inspection of the distribution, where some countries, such as the United States and Canada, have untracked curriculums. For this last variable we see no relationship at all, but after including the vocational index, the model is completely reversed with countries with tracked curriculum having about 0.6 more standard deviations in inequality that countries with no tracking.

@@

The biggest reductions are for both Hungary and Germany, with 46\% and 36\% respectively. France, which had a particularly steep increase in the last 15 years, would see a reduction of the 90/10 SES gap of 31\% and Austria, with a very similar tracking structure as the German case, would see their achievement gap reduced by 20\%. The very surprising result is the Dutch case, for which a switch from the 'worst' tracking setup to the 'best' case would result in an increase of the gap of 19\%.

This last result could mean two things. First and foremost, the Dutch educational system has built other mechanisms that, despite the tracking system, help children from low social origins escalate upwards. This explanation is not very trivial considering that in most mobility research, the Netherlands stands out as a very equal country relative to other countries with a similar tracking structure, like Germany \citep{shavit1993, esping2012}. Another possible explanation, and something which needs to assessed, is whether our model is predicting accurately. If we look at the plot from above, we see that for some countries, the model is a bit off from the actual level of inequality. Moreover, although we see that if all countries switched to the 'ideal' tracking setup we would see a vast reduction in inequality, we also see that each of these predictions is followed by a wide uncertainty interval.

@@

The exact mechanisms through which the tracking setup is explaining these inequalities is still very speculative. We could argue that tracking might've introduced the inequality once it was implemented. Another possibility is that tracking, although not changing, is currently exacerbating it every year. Another explanation, which we find strong evidence for, is that the tracking setup interacts very closely with the vocational setup. These explanations are very speculative and further research should carry the task of looking for a more concrete explanation.

@@

The literature on achievement inequality has recently started studying the evolution of the achievement gap and it has uncovered a wide variety of results. The main finding is that the U.S gap has been widening over time by about 40-50\% from 1970 until the 2000. Others find that this recent trend has reversed, at least for an age-specific group since the 2000's. From an international perspective, recent studies have shown that there is wide variability in this trend, with some countries showing decreases while others increases. This paper looks to investigate the magnitude of these gaps across selected countries, and attempts to explain why they are happening.

We start off by looking at the evolution of the achievement gap and document the same variation for the 15-year old achievement gap across countries. The United States has been closing its 90/10 gap by about 50\% with a wide interval going down to nearly 35\%. This reduction is also present in Germany, quite shockingly a country with a very institutionalized tracking system. Other countries such as France and Austria experience the contrary with persistent increases in the achievement gap. Moreover, once we disaggregate these patterns into a more fine-grained analysis, we discover that the dynamics are very different. The U.S gap is shrinking because both the 90th and 50th SES groups are actually going down while the bottom group is coming up. These results are evident because the 90/10 and 50/10 gap are shrinking at a very rapid rate, whereas the 90/50 gap is flat in the same 15 year interval. Germany, which has a very similar reduction, portrays a different explanation. All three achievement gaps are shrinking suggesting a more uniform reduction. On the other hand, for countries which are experiencing a widening of the gap, we see that it's happening at all instances of the distribution; all SES groups show a very steep increase.

@@

We complete the analysis by studying what explains the evolution of this inequality and we find that the degree of tracking of a country seems to be a possible explanation, but it should not be studied without factoring in the degree of vocational enrollment. We find that if a country has a low level of tracking, vocational enrollment will not help decrease the achievement gap. Conversely, once tracking enters the picture, vocational enrollment can help unease the burden of inequality and reduce the achievement gap very strongly. In a similar vein, we use this model to perform a simulation for each country: if every country switched to little or no tracking, what would be the resulting reduction in the 90/10 achievement gap? We find that if all countries switched, we would find that on average countries would see their gap shrink by about 20\%, with countries like Hungary and Germany experiencing reductions of 46\% and 36\% respectively. We contrast the former simulation with another counterfactual scenario: what would be the average gap increase if all countries switched to a system where tracking predominates? We record an average widening of 72\%, something that suggests that increasing tracking is more than three times worse than reducing it.

We conclude by showing that tracking and vocational enrollment are playing an important role in explaining the evolution of the achievement gap. We find that these results are very suggestive that institutional features, more specifically, stratified curriculums, might be playing a very sizable role in the promotion of inequality. After having ran a battery of tests and different empirical strategies, the results are consistent and point out that we should pay attention to the source of the achievement gaps and study whether particular SES groups are contributing more to the gap than others. Aside from the results, this paper also revealed some puzzling questions that we leave up to future research to study.
